Bayesian exponential random graph modeling of KWL Burial network

### 2020-06-24 analysis/paper/paper.Rmd
expand and edit method section

<!--chronology-->
We assigned burials to the pre-European period, European and post-European period for network comparison according to depth, radiocarbon dates, diagnostic materials, stratigraphic data, and previous studies. Since the sedimentary characteristics show vertical and lateral continuity cross units of AD section, depth is used as the first criterion for burial chronology. The chronology is then evaluated by the burials with time marker, light gray glazed jars, also called An-ping jars. The jars were believed to be produced in Southeast China and circulated in Southeast Asia from the late 16th to 18th century, but are often found at sites with the European occupation in the early 17th century in Taiwan [@Berrocal2018; @Hsieh1995; @Ketel2011]. In addition to the diagnostic jars, three radiocarbon dates of charcoal associated with three different burials are used as another evidence to evaluate the chronology [@Chen2007]. Some burials that were heavily disturbed by modern construction are excluded to avoid misinterpreting due to incomplete data. With the information of burials, including quantity of grave goods, types of grave goods, burial identities, we can reconstruct social relations between individuals by linking them according to similar variables. This is based on the principle that prestige materials usually relate to a representation of social status, where the same economic or social class, tend to use similar goods as a status indicators [citations]. Therefore, we build network where burials represent actors (nodes in the network) that are linked when they have th same prestige goods in common. The prestige goods we identified including agate beads, gold-foil beads, and imported Chinese porcelains, according to previous burials analysis [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011] and the description in historical records [@Borao2009; @LiandWu2006].

@@

We analyzed archaeological burials from a Iron Age site, Kiwulan, in northeastern Taiwan based on data collected from published excavation report, and original fieldwork notes to generate networks. The data includes information of gender, age, burial rites, and types of burial goods. We do not include the burials in Chinese phase in our network analysis because of the small sample size. A network is a structure can be represented as a graph consisting of a set of nodes (actors) and edges (ties) connecting them [@Brughmans2018; @Robins2007]; in our case, burials are nodes linked by edges if any pair of burial has the prestige goods in common. This is based on the assumption that the prestige goods imply social practices of trade, exchange, and gifting that reflect social relations between individuals [@Coward2013, pp. 252]. Since burials tend to have multiple prestige goods in common, such as personal ornaments and imported porcelains, the network ties are weighted instead of binary variables, where the value 1 represents a tie between two nodes and the value 0 otherwise [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors.

Our purpose is to understand the underlying mechanisms that shape the relationships between people at Kiwulan, and the effect of foreign contact on these relationships. In other words, we explore which variables may have contributed to the observed economic inequality across burials using network analysis approaches, and any changes in network structure before and after the foreign contacts. We hypothesized that after the start of the European presence in the 17th century, a group of ambitious individuals competed with each other for access to trade with the Europeans due to the high value of trade goods in their society. The high value of trade goods had been identified by the historical documents written by the Europeans and the Han Chinese [@Borao2009; @LiandWu2006]. In addition to valuable foreign goods, the indirect colonialism in northeastern Taiwan may suggest the potentially manipulation of the European colonial image by local ambitious individuals to built their personal status and power. This led to the monopolization of valuable goods among a small numbers of individuals and the society gradually switched to a structure with social differentiation. If the social inequality gradually increased in Kiwulan, then we will observe it from the burials by analyzing the networks generated by the social processes. This is important to understand how an Indigenous society reacted to the presence of foreign imperial power.

@@

We use Bayesian inference on exponential random graph models (ERGMs) to quantify the relationships among burials and test our hypothesis. ERGMs are statistical models for building complex social networks where possible ties are assumed as random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. In ERGMs, ties form a small structure in a network called graph configuration that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity (relationship between two actors through a shared third actor), homophily (relationship between actors with similar attribute), or popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations are the structure or property of a network and can be expressed as network statistics. By specifying the forms of configurations, we can build a hypothesis-based model to generate a distribution of random networks. This distribution consists of a large number of possible networks that provides a statistical inference to an observed network [@Robins2007]. This helps to understand whether our observed network shows significantly more or less property of interest than the large set of random networks we modeled.

Table \@ref(tab:ergm-terms-table) lists the parameters we used in the model construction for burials with archaeological interpretations. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. We test our hypothesis by measuring reciprocity, transitivity, popularity, and homophily structure effects. We expect that if our hypothesis is a good match with the data then we should see less reciprocally, less transitivity, but higher popularity structure effects after the European presence. We assigned the burials into three classes, high, medium, and low, as an index of wealth, based on the quantity of burial goods. In addition to wealth, we also consider another two social variables, age and gender of the interred individual, serving as indicators for distinguishing ascribed or achieved status in a society [citation].

@@

We analyzed social networks using the R programming language with bergm package for Bayesian exponential random graph models [@Caimo2014]. Bergm uses ergm packages for simulation and model specification, and provides a framework for Bayesian parameter estimation on the result of exponential random graph models [@Hunter2008; @Morris2008; @Handcock2008]. We set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical way to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network.

@@

<!-- briefly discuss Chinese burials -->

>>>>

<!--chronology-->
We assigned burials to the pre-European period (n = 26), European and post-European period (n = 45) for network comparison according to depth, radiocarbon dates, diagnostic materials, stratigraphic data, and previous studies. Since the sedimentary characteristics show vertical and lateral continuity cross units of AD section, depth is used as the first criterion for burial chronology. The chronology is then evaluated by the burials with time marker, light gray glazed jars, also called An-ping jars. The jars were believed to be produced in Southeast China and circulated in Southeast Asia from the late 16th to 18th century, but are often found at sites with the European occupation in the early 17th century in Taiwan [@Berrocal2018; @Hsieh1995; @Ketel2011]. In addition to the diagnostic jars, three radiocarbon dates of charcoal associated with three different burials are used as another evidence to evaluate the chronology [@Chen2007]. Some burials that were heavily disturbed by modern construction are excluded to avoid misinterpreting due to incomplete data. With the information of burials, including quantity of grave goods, types of grave goods, burial identities, we can reconstruct social relations between individuals by linking them according to similar prestige goods. This is based on the assumption that the prestige goods imply social practices of trade, exchange, and gifting that reflect social relations between individuals [@Coward2013, pp. 252]. Also, prestige materials usually relate to a representation of social status, where the same economic or social class, tend to use similar goods as a status indicators [citations]. Therefore, we build network where burials represent actors (nodes in the network) that are linked when they have th same prestige goods in common. The prestige goods we identified including agate beads, gold-foil beads, imported Chinese porcelains, gold foils, and fish-shaped ornaments, according to previous research [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011] and the description in historical records [@Borao2009; @LiandWu2006].

@@

We analyzed archaeological burials from an Iron Age site, Kiwulan, in northeastern Taiwan based on data collected from published excavation report, and original fieldwork notes to generate networks. We do not include the burials in Chinese phase in our network analysis because of the small sample size. A network is a structure can be represented as a graph consisting of a set of nodes (actors) and edges (ties) connecting them [@Brughmans2018; @Robins2007]. In our case, burials are nodes linked by edges if any pair of burial has the prestige goods in common. Due to gold-foil beads and agate beads commonly found across burials with substantial differences in counts, we assigned those prestige goods into three levels based on the distribution of their counts to indicate quantity, high, medium, and low. That means if burial 1 and burial 2 both have high quantity of agate beads, then there will be a tie connecting them. For prestige goods not commonly found across burials, including imported porcelains, gold foils, and fish-shaped ornaments, we assign a tie between burials when they possess the good in common. Node attributes here include osteological data and mortuary practices, such as age, gender, burial ritual, and wealth level. Burial ritual is identified by locally made ceramics on the top of grave that suggests the presence of funeral feasting. Wealth level is an important attribute for economical inequality that we assigned the burials into three classes, high, medium, and low, as an index of wealth, based on the quantity of burial goods. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors.

We hypothesized that after the start of the European presence in the 17th century, a group of ambitious individuals competed with each other for access to trade with the Europeans due to the high value of trade goods in their society. The high value of trade goods had been identified by the historical documents written by the Europeans and the Han Chinese [@Borao2009; @LiandWu2006]. In addition to valuable foreign goods, the indirect colonialism in northeastern Taiwan may suggest the potentially manipulation of the European colonial image by local ambitious individuals for building their personal status and power. This led to the monopolization of valuable goods among a small numbers of individuals and the society gradually switched to a structure with more social inequality. If the social inequality gradually increased in Kiwulan, then we will observe it from the burials by analyzing the networks generated by those social processes. Network analysis provides a way to understand the underlying mechanisms that shape the relationships between people at Kiwulan, and the effect of foreign contact on these relationships. Our purpose is to identify changes in network structure of burials before and after the foreign contacts, and explore which variables may have contributed to the observed economic inequality across burials using network analysis approaches. This is important to understand how an Indigenous society reacted to the presence of foreign imperial power.

@@

We use Bayesian inference on exponential random graph models (ERGMs) to quantify the relationships among burials and test our hypothesis. ERGMs are statistical models for building complex social networks where possible ties are assumed as random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. In ERGMs, ties form a small structure in a network called a graph configuration that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), or popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. By specifying the forms of configurations, we can build a hypothesis-based model to generate a distribution of random networks. Such distribution consists of a large number of possible networks that provides a statistical inference for an observed network [@Robins2007]. This helps to understand whether an observed network shows significantly more or less property of interest than the random networks we modeled.

Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with archaeological interpretations. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. We test our hypothesis by measuring reciprocity, transitivity, popularity, and homophily structure effects. We expect that if our hypothesis is a good match with the data then we should see less reciprocally, less transitivity, but higher popularity structure effects after the European presence. This structure infers the presence of a few individuals as centers having more relations with others with a decrease in small and multiple groups, which hints the emergence of social inequality. We also consider homophily effects based on node attributes, including age, gender, ritual, and the degree of wealth, to explore the correlation between the tendency of connected nodes and the network ties overall. This can imply the major effect that contributes to the network that we observed and give an insight into the underlying mechanism of social inequality if any.

@@

We analyzed social networks using the R programming language with bergm package for Bayesian exponential random graph models [@Caimo2014]. Bergm provides a framework for Bayesian parameter estimation on the result of exponential random graph models, which depends on ergm packages for simulation and model specification [@Hunter2008; @Morris2008; @Handcock2008]. Despite exponential random graph models can better describe a network by capturing the dependency structure, it is difficult to estimate model parameters and interpret the result due to the intractable normalizing constant and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a step to make probability distribution to integrate to one, which becomes hard to compute for larger set of networks [@Caimo2017bayesian]. Another problem is the degeneracy that means the probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. To overcome these limitations, Bayesian model applies Markov chain Monte Carlo (MCMC) simulation with the exchange algorithm, which solves computational problems in ERGMs [@Caimo2011].

MCMC simulation can estimate the posterior distribution of interest by of random sampling without the need of normal distribution for the prior and likelihood. MCMC is a random probability distribution that describes a sequence of events or transitions (chain) from one state to another state following probabilistic rules. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and than switch to a pair of actors at random within that chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical way to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network.

@@

<!-- briefly discuss Chinese burials -->
Age and gender attributes serve as indicators for distinguishing ascribed or achieved status in a society [citation].
